On the analytic structure of the linear response susceptibility in open systems

### 2018-03-21 AnalyticStructure_main.tex
update

%% Language and font encodings
\usepackage[english]{babel}
%\usepackage[utf8x]{inputenc}
%\usepackage[T1]{fontenc}

%% Useful packages
\usepackage{graphicx}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{xifthen}

@@

State something about the assessment of physical reliability for a ground state calculation. Some comment on the setup independence of the results. Introduce the correspondence
between localization properties and observable behavior or physicality. State the, at the present days, similar consideration for the results of linear response are still 
missing...

The basic motivation at the basis of the analysis discussed in this paper is to address to following question: which are the requirement that a computational setup has to
satisfy in order to be able to provide physical results for a linear response calculation? Here with physical we meant that they depend only on the features of the system under
inspection and on the perturbation acting on it. Physical results are unaffected by computational parameters, like the dimension of the computational domain or the nature of
the basis set used to perform the calculations.

To answer to this question we investigate the properties of the susceptibility, which represents the fundamental object in the contest of linear response.

@@

Linear response can be effectively applied to the analysis of the electronic excitations in molecular and condensed-matter systems. In this framework one consider a system 
described by the Hamiltonian $\hnot$ and assume that the ground-state density $\dmnot$ has been provided. Linear response formalism allows us to evaluate the modification 
of the expectation value of a generic observable induced by a time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed 
through the evaluation of the \emph{linear response functional}:
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
where the operator $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} 
under inspection. The modification of the ground-state density induced by the perturbation is expressed by the \emph{response density operator} $\dm'(\omega)$ that, in the linear 
response regime, is a linear function of the perturbing field $\op\Phi(\omega)$. For this reason it is possible to define the \emph{susceptibility} superoperator, as follows:
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \sop{{\cal R}}(\omega) \op\Phi(\omega) \;. 
\ee
Susceptibility describes an intrinsic feature of the unperturbed system and determines its response to \emph{any} perturbations, in the linear response regime.

The analysis presented in this section is devoted to a critical review of the fundamental equations which define the response density operator in the linear response regime.    
The aim of this process is to....

%the construction of two different representations of the response density operator...

\subsection{Fluctuation states. A (de-)localization argument for the response density}

@@

The interesting aspect of equations \eqref{fluctuationStateEqMotion1} in the perspective of the present analysis is that they allow us to present a \emph{localization argument} 
for the fluctuation states. To clarify this statement we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and solve \eqref{fluctuationStateEqMotion1} 
in the form of a self-consistent inhomogeneous Helmholtz equation by writing:
\be
\ket{f_p(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s(\omega)} \;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\omega) = (\omega-\op T)^{-1}$, % with $k^2_p(\omega) = (\omega+\eps_p)$,
and the \emph{source term} is given by:
\be
\ket{s(\omega)} =  \op V\ket{f_p(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
For open boundary conditions in three dimensions, the kernel of $\op G_H(\omega)$ in the coordinate representation reads: 
% \be\lb{HelmholtzKernelDef1}
% \bra{\r}\op G_H(\omega)\ket{\r'} = \frac{1}{4\pi} \left\{ 
% \begin{matrix}
% \frac{e^{-2k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega \leq |\eps_p| \\ 
% \frac{e^{\ii k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega > |\eps_p|
% \end{matrix}
% \right. \;,
% \ee
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\omega) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{2k|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; k^2 = \omega  < 0\; \\ 
\frac{e^{2\ii k |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; k^2 = \omega > 0
\end{cases} \;,
\ee
and the $\r$-representation of the fluctuation states $f_p(\omega,\r)$ is expressed through a convolution integral between the kernel \eqref{HelmholtzKernelDef1} and
$\brket{\r'}{s(\omega)}$.
An inspection of the structure of \eqref{HelmholtzKernelDef1} evidences that, for each value of $p$, there exist a threshold value $\omega = |\eps_p|$, 
which govern the localization properties of the corresponding fluctuation state. Indeed, for $\omega > |\eps_p|$,  $\op G_H(\omega)$ behaves as a spherical wave and its
convolution with the source term always give raises to a $f_p(\omega,\r)$ delocalized in the whole accessible space. On the contrary, for values of $\omega$ below the threshold,
the kernel contains an exponential damping factor and $f_p(\omega,\r)$ exhibit a bound-state behavior in the long range.

@@

\subsection{Excitation operators. A spectral decomposition representation for the response density}

The analysis carried out starting from the representation of the linear response functional in terms of fluctuation states, has evidenced some general properties of the linear response
for open systems. In particular, by analyzing the asymptotic behavior of fluctuation states in long range limit, we have showed that they exhibit a bound-state character in the regime  
$\omega<|\eps_h|$. consequently we have argued that the linear response functional, in the same $\omega$ range, provides a setup-independent result with a modest computational effort.

>>>>

% Language and font encodings
\usepackage[english]{babel}
%\usepackage[utf8x]{inputenc}
%\usepackage[T1]{fontenc}

% Useful packages
\usepackage{graphicx}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{xifthen}

@@

The subject of the present investigation concerns with the study of isolated molecular-like systems subjected to a time-dependent perturbation. % in the linear response regime.
This problem can be tackled within the contest of the time-dependent density functional theory (TD-DFT)\cite{runge1984,onida2002}, that allows one to establish the time-dependent 
Kohn-Sham (KS) equations which determine the evolution of the one-particle properties of the system with respect to time.

We focus our attention to the linear response regime (LR) in which the theory can be conveniently expressed in the frequency domain and the time-dependent KS equation are reformulated 
as a quantum Liouville equation for the response density operator. Different computational approaches have been developed in literature\cite{baroni2008,hubener2014,brabec2015}, principally 
with the aim of producing efficient algorithms for the evaluation of the physical observables. These methods differ both for the procedure adopted to solve the Liouville equation 
(explicit diagonalization, self-consistent Sternheimer equation, Lanczos chain-like procedure, etc.) and both for the choice of the basis set employed to perform the calculations. 
Nonetheless, in all the cases one is a dealing with a finite basis set that, unavoidably, gives rise to a limited computational domain. It is then meaningful to try to understand under 
which extent the computational-induced truncation of the natural open boundaries condition could produce observable effects on the physical quantities computed in the LR.

A preliminary understanding of this problem can be achieved thanks to the results reported in\cite{boffi2016}. The authors of this paper discuss the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, and the independence of its properties (\emph{e.g.} energy) 
with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

In this paper we aim to reach a deeper understanding of the possible effects of a finite computational basis set on the LR regime in open systems. To achieve this task we try to
reformulate the relation between localization properties and basis independence for the main actors of the LR formalism. We present a discussion based on a double level analysis. First, 
we propose a localization argument directly based on the inspection of the response density operator. Then, we move to a deeper level and investigate the properties of the susceptibility 
functional, which represents the fundamental object in the contest of the linear response. Pursuing this approach we are naturally led to investigate the localization properties of excitation 
operators, that in turn emerge as the resonant channel of the spectral decomposition of the susceptibility. In this way we establish a direct relation between the concept of basis independence 
and the analytic structure of the linear response susceptibility.

% First of all it is useful to discuss the property of the ground state under this perspective. 
% Indeed, it is well known that the stability of the ground state and the localized nature of a molecular-like system ensures that the KS description of occupied states is represented 
% by a discrete set of wave functions that approaches to zero far apart from a compact region around the system's atoms. This behavior indicates that a (finite) basis set will provide 
% a reliable description of the ground state properties if it is able to sample a computational domain which contains the natural sizes of the occupied orbitals\cite{boffi2016}.
% The same arguments cannot be rigidly extended to the case in which an external time-dependent field starts acting on the system. Indeed, the perturbation drives the system out 
% from the equilibrium and the electrons sample a wider region of the state space, that will contains orbitals of both negative (bound) and positive (unbound) energies.

@@

In this section we review the fundamental equations which govern the behavior of an open quantum system in the linear response regime. Our analysis starts with a brief description of 
the notation adopted in the paper. Then we move on the analysis of the response density and of the excitation operators, with the aim to establish their localization character.

%The analysis presented in this section is devoted to a critical review of the fundamental equations which define the response density operator in the linear response regime.

Linear response can be effectively applied to the analysis of the electronic excitations in molecular systems. In this framework one consider a system 
described by the Hamiltonian $\hnot$ and assume that the ground-state density $\dmnot$ has been provided. Linear response formalism allows us to evaluate the modification 
of the expectation value of a generic observable induced by a time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed 
through the evaluation of the \emph{linear response functional}:
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
where the operator $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} 
under inspection. The modification of the ground-state density induced by the perturbation is expressed by the \emph{response density operator} $\dm'(\omega)$ that, in the linear 
response regime, is a linear function of the perturbing field $\op\Phi(\omega)$. For this reason it is possible to define the \emph{susceptibility} superoperator, as follows:
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \sop{{\cal R}}(\omega) \op\Phi(\omega) \;. 
\ee
Susceptibility describes an intrinsic feature of the unperturbed system and determines its response to \emph{any} perturbations, in the linear response regime.

\subsection{Fluctuation states. A localization argument for the response density}

@@

The interesting aspect of equations \eqref{fluctuationStateEqMotion1} in the perspective of the present analysis is that they allow us to present a \emph{localization argument} 
for the fluctuation states. To clarify this statement we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and solve \eqref{fluctuationStateEqMotion1} 
in the form of a self-consistent inhomogeneous Helmholtz equation by writing:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s(\omega)} \;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\omega) = (\omega-\op T)^{-1}$, % with $k^2_p(\omega) = (\omega+\eps_p)$,
and the \emph{source term} is given by:
\be
\ket{s(\omega)} =  \op V\ket{f_p(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
For open boundary conditions in three dimensions, the kernel of $\op G_H(\omega)$ in the coordinate representation reads: 
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\omega) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{-\alpha|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; \omega  < 0\; \\ 
\frac{e^{\ii \alpha |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; \omega \geq 0
\end{cases} \;,
\ee
where $\alpha = \sqrt{2|\omega|}$.

The localization character of fluctuation states can be determined by looking at the long range behavior of \eqref{rhoPrimeFluctuationStateDef2}, projected in the $\r$-representation.
An inspection of the structure of \eqref{HelmholtzKernelDef1} evidences that, for each value of $p$, there exist a threshold value $\omega = |\eps_p|$, which govern the localization 
properties of the corresponding fluctuation state. Indeed, for $\omega > |\eps_p|$,  $\op G_H(\omega)$ behaves as a spherical wave and its convolution with the source term  gives 
raise to a $f_p(\omega,\r)$ delocalized in the whole accessible space. On the contrary, for values of $\omega$ below the threshold,
the kernel contains an exponential damping factor and $f_p(\omega,\r)$ exhibit a bound-state behavior in the long range.

@@

\subsection{Excitation operators. Spectral decomposition of the susceptibility functional}

The analysis carried out starting from the representation of the linear response functional in terms of fluctuation states has evidenced some general properties of the linear response
for open systems. In particular, by analyzing the asymptotic behavior of fluctuation states in long range limit, we have showed that they exhibit a bound-state character in the regime  
$\omega<|\eps_h|$. Consequently we have argued that the linear response functional, in the same $\omega$ range, provides a setup-independent result with a modest computational effort.
